---
title: "Stat 4620 Final Project Initial EDA"
author: "Nicolas Brizzi"
date: "2022-11-28"
output: html_document
---

```{r}
library(dplyr)
library(magrittr)
library(tidyr)
library(GGally)
```

```{r}
train = read.csv("/Users/Nick/Downloads/Ames/train.csv")
```

## EDA 

Frequency of data types in the Ames frame:
```{r}
table(sapply(train, class))
```
About half and half - slightly more character data types. 

Quick data summaries.
```{r}
summary(train)
```

How much data is missing?
```{r}
sum(is.na(train))
```
6965 different values are NA

Let's replace those NA values with 0 or None depending on the variable type. 
```{r}
for (i in 1:length(names(train))) {
  if(class(train[[i]]) == "character") {
    train[[i]] <- train[[i]] %>% replace_na("None")
  }
  else {
    train[[i]] <- train[[i]] %>% replace_na(0)
  }
}
```

```{r}
sum(is.na(train))
```

### Seeing if there is any linear relationships between some of the numeric predictors that we can pick up on. 

Looking
```{r}
train %>%
  select(contains("Garage")) %>%
  select_if(is.numeric) %>%
  ggpairs()
```
It looks like 'GarageCars' and 'GarageArea' have a very high linear correlation, which is to be expected. 
The other garage variables have moderately strong linear correlations amongst themselves. 

Variables that contain "Overall"
```{r}
train %>%
  select(contains("Overall")) %>%
  select_if(is.numeric) %>%
  ggpairs()
```

How much can Overall Quality tell us about the house sale price?
```{r}
lm(SalePrice ~ OverallQual, data = train) %>%
  summary()
```
R-squared of ~.63, decently strong correlation for just one variable. This most likely will be included in the model that we end up on. 


Variables that contain "SF" or Square Feet
```{r}
train %>%
  select(contains("SF")) %>%
  select_if(is.numeric) %>%
  ggpairs()
```
There's a lot going on but there are some strong correlations here. 
- '1stFlrSF' and 'TotalBsmtSF' have a correlation of .820
- 'BsmtFinSF1' and 'TotalBsmtSF' have a correlation of .522

All other correlations are below .5. 

## Initial Considerations

So far there is a lot of character variables that can be interpreted as ordinal, and a lot of numeric variables. These variables are good for techniques that have a lot of quantitative variables that have complex, and perhaps non-linear, relationships among each other. It can potentially eliminate a lot of factors and retain an accurate model. 

It will be important to decide on what factors to convert into numeric values, and what variables to exclude from the model as a whole. 
